---
layout: post
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
---
layout: post
title: Is R Fast Enough? - Part 3 - "Fibonacci"
author: Eliot McIntire
date: May 6, 2015
tags: [R]
comments: true
---

In part 3 of this series on benchmarking `R`, it will be quick. Just the Fibonacci series. This is one that R gets a really bad reputation about.  Because it is an iterative function, it can't be vectorized, which is the usual way that we make R programs fast. Doing explicit loops in `R` is thought to be slow. So lets look...

Similarly, [julialang.org](http://julialang.org) showed that Fibonacci series was, again, particularly bad in `R`. We, again, felt that this was a case of poor `R` coding, or more accurately, missing the point of whether `R` was capable or not.  

### Fibonacci Series

We run 2 versions of `C++`, 2 versions of `R` that we build here, and 1 version within the `numbers` package from the open source R community.

```{r load_cpp_functions, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE}
library(rbenchmark)
library(Rcpp)
library(numbers)
library(microbenchmark)
```

```{r RcppFunctions1, cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE}
cppFunction('int fibCpp2(const int x) {
    if (x == 0 || x == 1) return(x);
    return (fibCpp2(x - 1)) + fibCpp2(x - 2);
}')

cppFunction('int fibCpp1(int n) {
    return n < 2 ? n : fibCpp1(n-1) + fibCpp1(n-2);
}')
all.equalV = function(...) {
  vals <- list(...)
  all(sapply(vals[-1], function(x) all.equal(vals[[1]], x)))
}

```


```{r fib, eval=TRUE, echo=TRUE, cache=TRUE}

# Define two R functions
fibR1 = function(n) {
    fib <- numeric(n)
    fib[1:2] <- c(1, 2)
    for (k in 3:n) {
        fib[k] <- fib[k - 1] + fib[k - 2]
    } 
    return(fib)
}
fibR2 = function(n) {
     if (n < 2) {
         return(n)
     } else {
         return(fibR2(n-1) + fibR2(n-2))
     }
}

N = 20L
mbFib <- microbenchmark(times=200L, 
                    a <- fibonacci(N+1, TRUE)[N+1], 
                    b <- fibCpp1(N+1), 
                    d <- fibCpp2(N+1), 
                    e <- fibR1(N)[N], 
                    f <- fibR2(N+1))
summary(mbFib)[c(1,2,5,7)]
all.equalV(a,b,d, e, f)
```

Here, one of the two native R implementations is **`r round(summary(mbFib)[[4]][5]/summary(mbFib)[[4]][4])`x faster** by pre-allocating the output vector size. The fibonacci function in the package `numbers` was `r round(summary(mbFib)[[4]][1]/summary(mbFib)[[4]][3],2)`x slower than the faster `R` function because it has error checking. ***The native C++ version was `r round(summary(mbFib)[[4]][2]/summary(mbFib)[[4]][4],2)`x slower***. 

#### Take home points:

1. *Pre-allocate* vectors. This is standard in other languages, yet it is not done in many basic tests of R code.

`R` certainly held its own, again, compared to simple `C++` functions precompiled using `Rcpp` package. In this case, we used native `R` with pre-allocation, and it was faster than the fastest `C++` version. And of course, there was a very slow way to do things in `R` as well. The function within the package `numbers` was very fast and had nice checking within the function that is likely worth the overhead in many cases. 

#### Conclusion

*YES! `R` is more than fast enough*.  

#### Next time - Loops

How to make loops in R faster, for those times that you can't make code vectorized.

--------------------

#### Functions used

The C++ functions that were used are:

```{r FunctionsUsed, cache=FALSE, eval=FALSE, echo=TRUE, message=FALSE}
cppFunction('int fibCpp2(const int x) {
    if (x == 0 || x == 1) return(x);
    return (fibCpp2(x - 1)) + fibCpp2(x - 2);
}')

cppFunction('int fibCpp1(int n) {
    return n < 2 ? n : fibCpp1(n-1) + fibCpp1(n-2);
}')

all.equalV = function(...) {
  vals <- list(...)
  all(sapply(vals[-1], function(x) all.equal(vals[[1]], x)))
}

```

#### System used:
Tests were done on an HP Z400, Xeon 3.33 GHz processor, running Windows 7 Enterprise, using:

```{r session_info, eval=TRUE, echo=FALSE}
sessionInfo()
```
