---
layout: post
output:
  html_document:
    keep_md: yes
---
---
layout: post
title: Is `R` Fast Enough? - Part 2 - "Sorting"
author: Eliot McIntire
date: April 28, 2015
tags: [R, benchmark]
comments: true
---

In part 2 of this series on benchmarking `R`, we'll explore sorting. This has been a topic on numerous blogs, discussions and posts around the internet, including here: [r-blogger post](https://www.r-bloggers.com/quicksort-speed-just-in-time-compiling-and-vectorizing/). Similarly, [julialang.org](https://julialang.org) showed that sorting was particularly bad in `R`. We, again, felt that this was a case of poor `R` coding, or more accurately, missing the point of whether `R` was capable or not. Another [example here](https://gallery.rcpp.org/articles/sorting/) compares `R` sorting with standard library of C++, called from `R`. 

In all cases, we felt that one of the points of using `R` is that there are concise ways of doing things **because the open source community has brought them to `R`**. So lets take advantage of that! We will cover both real number sorting and integer sorting. 

### Sorting

```{r load_cpp_functions, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE}
library(rbenchmark)
library(dplyr)
library(Rcpp)
library(data.table)
```

```{r RcppFunctions1, cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE}
cppFunction('NumericVector stl_sort(NumericVector x) {
   NumericVector y = clone(x);
   std::sort(y.begin(), y.end());
   return y;
}')

```


This was in part inspired from a blog post by Wingfeet at https://www.r-bloggers.com/quicksort-speed-just-in-time-compiling-and-vectorizing/ which drew on benchmark tests here: https://julialang.org/ 
Essentially, that julia test was a benchmark to test the speed of Julia. It showed for the Quicksort, that `R` is 524x slower than C. Below is that version. But, there was no explicit comparison of how the base `R` sort would match with C, nor how any of the more recent packages with sorting capability fare against these procedural versions of low level languages. 

```{r sorting_fns, eval=TRUE, echo=FALSE, cache=FALSE}
qsort = function(a) {
    qsort_kernel = function(lo, hi) {
        i = lo
        j = hi
        while (i < hi) {
            pivot = a[floor((lo+hi)/2)]
            while (i <= j) {
                while (a[i] < pivot) i = i + 1
                while (a[j] > pivot) j = j - 1
                if (i <= j) {
                    t = a[i]
                    a[i] <<- a[j]
                    a[j] <<- t
                    i = i + 1;
                    j = j - 1;
                }
            }
            if (lo < j) qsort_kernel(lo, j)
            lo = i
            j = hi
        }
    }
    qsort_kernel(1, length(a))
    return(a)
}

wfqs1 <- function(x) {
  if (length(x)<2) return(x)
  pivot <- x[sample(length(x),1)]
  c(wfqs1(x[x<pivot]),x[x==pivot],wfqs1(x[x>pivot]))
}

wfqsx = function(a) {
  qsort_kernel = function(lo, hi) {
    if(lo>=hi) return()
    if(hi-lo==1) {
      if(a[lo]>a[hi]) {
        t <- a[lo]
        a[lo] <<- a[hi]
        a[hi] <<- t
      }
      return()
    }
    goUp <- a[(lo+1):hi]>a[lo]
    up <- which(goUp)
    do <- which(!goUp)
    pivottarget <- lo+length(do)
    up <- up[up<=length(do)]+lo
    do <- do[do>length(do)]+lo
    t <- a[do]
    a[do] <<- a[up]
    a[up] <<- t
    t <- a[pivottarget]
    a[pivottarget] <<- a[lo]
    a[lo] <<- t  
    qsort_kernel(lo,pivottarget-1)
    qsort_kernel(pivottarget+1, hi)
  }
  qsort_kernel(1, length(a))
  return(a)
}

all.equalV = function(...) {
  vals <- list(...)
  all(sapply(vals[-1], function(x) all.equal(vals[[1]], x)))
}
```


#### Real number sorting

```{r sorting, eval=TRUE, echo=TRUE, cache=FALSE}
x = runif(1e5)
xtbl <- tbl_df(data.frame(x=x))
(mbReal <- benchmark(
           a <- qsort(x), 
           d <- sort(x), 
           e <- sort(x, method="quick"),
           f <- .Internal(sort(x,decreasing = FALSE)),
           g <- data.table(x=x,key="x"), 
           h <- arrange(xtbl,x),
           i <- stl_sort(x),
           replications=25L, columns=c("test", "elapsed", "relative"),
           order="relative"))
all.equalV(a, d, e, f, g$x, h$x, i)
```

#### Integer sorting

```{r integer_sort, eval=TRUE, echo=TRUE, cache=FALSE}
x = sample(1e6,size = 1e5)
xtbl <- tbl_df(data.frame(x=x))
(mbInteger <- benchmark(
           a <- qsort(x), 
           d <- sort(x), 
           e <- sort(x, method="quick"), 
           f <- .Internal(sort(x,decreasing = FALSE)),
           g <- data.table(x=x,key="x"), h<-arrange(xtbl,x),
           i <- stl_sort(x),
           replications=25L, columns=c("test", "elapsed", "relative"),
           order="relative"))
all.equalV(a, d, e, f, g$x, h$x, i)

```

Both real numbers and integers can be sorted quickly with `R`. The slowest function is indeed the procedural `qsort` written in native `R` without any optimization. This was also the `qsort` that the Julia testers used. Our numbers match almost exactly those from the the table in julialang.org; however, here we also test the real world `R` usage that a normal `R` user would face (*i.e.*, we can all use `sort()`). We show that `R` competes quite favourably and regularly outperforms standard library of C++ (and Julia!, though that is not tested here explicitly). 

#### Take home points:

1. the basic `R` sorting functions are fast. The `sort(method="quick")` is about as fast as the standard `C++` library sort (`r round((1-1/(mbInteger$relative[grep(as.character(mbInteger$test), pattern="stl_sort")]/mbInteger$relative[grep(as.character(mbInteger$test), pattern="quick")]))*100)`% faster). 
2. using [data.table](https://cran.r-project.org/web/packages/data.table/index.html) on integers is `r round((1-1/(mbInteger$relative[grep(as.character(mbInteger$test), pattern="stl_sort")]/mbInteger$relative[grep(as.character(mbInteger$test), pattern="data.table")]))*100)`% faster than the `C++` standard library sort.

In real world situations, where we want to use the easiest, shortest code to produce fast, accurate results, `R` certainly holds its own compared to the standard `C++` library. But of course, there are many ways to do things in `R`. Some are much faster than others.

#### Conclusion

Using the `sort(method="quick")` and [`data.table`](https://cran.r-project.org/web/packages/data.table/index.html) sorting, we were able to sort a vector of real numbers ***`r round(mbReal$relative[grep(as.character(mbReal$test), pattern="qsort")]/mbReal$relative[grep(as.character(mbReal$test), pattern="data.table")])`x*** faster than a naive procedural coding (`qsort`) and ***`r round(mbInteger$relative[grep(as.character(mbInteger$test), pattern="qsort")]/mbInteger$relative[grep(as.character(mbInteger$test), pattern="data.table")])`x*** faster on a vector of integers. These put the `data.table` sort as fast as or substantially faster than C or Fortran or Julia's version of quicksort (based on timings on [julialang.org](https://julialang.org)).

*YES! `R` is more than fast enough*.  

#### Next time (really! I promised it last time...)

We will redo the Fibonacci series, a common low level benchmarking test that [shows `R` to be slow](https://julialang.org).  But it turns out to be a case of bad coding...


--------------------

#### Functions used

The C++ functions that were used are:


```{r FunctionsUsed, cache=FALSE, eval=FALSE, echo=TRUE, message=FALSE}
cppFunction('NumericVector stl_sort(NumericVector x) {
   NumericVector y = clone(x);
   std::sort(y.begin(), y.end());
   return y;
}')
qsort = function(a) {
    qsort_kernel = function(lo, hi) {
        i = lo
        j = hi
        while (i < hi) {
            pivot = a[floor((lo+hi)/2)]
            while (i <= j) {
                while (a[i] < pivot) i = i + 1
                while (a[j] > pivot) j = j - 1
                if (i <= j) {
                    t = a[i]
                    a[i] <<- a[j]
                    a[j] <<- t
                    i = i + 1;
                    j = j - 1;
                }
            }
            if (lo < j) qsort_kernel(lo, j)
            lo = i
            j = hi
        }
    }
    qsort_kernel(1, length(a))
    return(a)
}

all.equalV = function(...) {
  vals <- list(...)
  all(sapply(vals[-1], function(x) all.equal(vals[[1]], x)))
}

```

#### System used:
Tests were done on an HP Z400, Xeon 3.33 GHz processor, running Windows 7 Enterprise, using:

```{r session_info, eval=TRUE, echo=FALSE}
sessionInfo()
```
