---
author: "Eliot McIntire and Ian Eddy"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  tidy: false
warning: false
message: false
error: false
cache: true
---

# Forest Succession and Wildfire using `setupProject` {#sec-scfmEg}

```{r setup}
#| include: false
#| cache: false

## truncate text output
hook_output <- knitr::knit_hooks$get("output")

knitr::knit_hooks$set(output = function(x, options) {
if (!is.null(n <- options$outlines)) {
x <- xfun::split_lines(x)
if (length(x) > n) {
# truncate the output
x <- c(head(x, n), "....\n")
}
x <- paste(x, collapse = "\n")
}
hook_output(x, options)
})

options(repos = list(CRAN = "http://cran.r-project.org"))

if (Sys.getenv("USERNAME") == "GHA") {
googledrive::drive_auth(path = Sys.getenv("GOOGLE_APP_CREDENTIALS"))
}

qmdPath <- normalizePath(getwd(), winslash = "/")

if (!interactive()) {
dir.create("~/tutos/castorExample", recursive = TRUE)
knitr::opts_knit$set(root.dir = "~/tutos/castorExample")
}
```


We have developed an R version of LANDIS-II's Biomass Succession Extension, comprised
of 2 modules: `Biomass_core` and `Biomass_regeneration`. We can combine these with
a fire model, such as `scfm` which is a collaborative effort lead by Steve Cumming at 
Laval, University (thus the name: **S**teve **C**umming **F**ire **M**odel).

As with other examples, we also include all the parameterization modules here, with 
`Biomass_borealDataPrep` for the `Biomass_***` modules and `scfmLandcoverInit`, 
`scfmRegime` and `scfmDriver` for the `scfm` family.

This makes for a 10 module project. Using `setupProject`, we can readily run these
from their canonical "i.e., current, up to date, open and available" versions:

**Time to run this** is between 1 and 2 hours.

```{r scfm,eval=FALSE}
repos <- unique(c("predictiveecology.r-universe.dev", getOption("repos")))
options(repos = repos)

# Require::Require(c("SpaDES.project (HEAD)"))
Require::Require(c("PredictiveEcology/SpaDES.project@development (HEAD)")) 

out <- SpaDES.project::setupProject(
  updateRprofile = TRUE,
  name = "scfm_example_fresh",
  useGit = FALSE,
  modules = c("PredictiveEcology/Biomass_borealDataPrep@development",
              "PredictiveEcology/Biomass_core@development",
              "PredictiveEcology/Biomass_regeneration@development",
              file.path("PredictiveEcology/scfm@development/modules",
                        c("scfmLandcoverInit", "scfmRegime", "scfmDriver",
                          "scfmIgnition", "scfmEscape", "scfmSpread",
                          "scfmDiagnostics"))
  ),
  params = list(
    .globals = list(.studyAreaName = "scfm_example",
                    dataYear = 2011, #will get kNN 2011 data, and NTEMS 2011 landcover
                    sppEquivCol = "LandR"),
    scfmDriver = list(targetN = 1000, #default is 4000 - higher targetN adds time + precision
                      ### targetN would ideally be minimum 2000 - mean fire size estimates will be bad with 1000
                      .useParallelFireRegimePolys = TRUE) #assumes parallelization is an otpion

  ),
  options = list(spades.allowInitDuringSimInit = TRUE),
  #                spades.moduleCodeChecks = FALSE,
  #                reproducible.shapefileRead = "terra::vect",
  #                spades.recoveryMode = 1
  # ),
  packages = c('RCurl', 'XML', 'snow'), # need for some downloading; if omitted, a message tells user
  times = list(start = 2011, end = 2061),
  sppEquiv = LandR::sppEquivalencies_CA[KNN %in% c("Popu_Tre", "Betu_Pap",
                                                   "Pice_Gla", "Pice_Mar",
                                                   "Pinu_Con", "Pinu_Ban")],
  studyArea = {
    targetCRS <- paste("+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +x_0=0 +y_0=0",
                       "+datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0")
    sa <- terra::vect(cbind(-1209980, 7586865), crs = targetCRS)
    sa <- LandR::randomStudyArea(center = sa, size = 10000 * 250 * 30000, seed = 1002)
    sa <- sf::st_as_sf(sa)
  },
  studyAreaLarge = {
    sf::st_buffer(studyArea, 20000)
  },
  rasterToMatchLarge = {
    rtml<- terra::rast(terra::ext(studyAreaLarge), res = c(250, 250))
    terra::crs(rtml) <- terra::crs(studyAreaLarge)
    rtml[] <- 1
    rtml <- terra::mask(rtml, studyAreaLarge)
  },
  rasterToMatch = {
    rtm <- terra::crop(rasterToMatchLarge, studyArea)
    rtm <- terra::mask(rtm, studyArea)
  }
)


#this must be done outside of setupProject (temporarily)
#alternatively
outSim <- do.call(SpaDES.core::simInitAndSpades, out)
```
