---
title: "LandR workflow with `setupProject`"
author: "Ceres Barros"
date: last-modified
editor_options:
  chunk_output_type: console
bibliography: references/references.bib
---

```{r setup}
#| include: false
#| cache: false

knitr::opts_chunk$set(tidy = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE)

## truncate text output
hook_output <- knitr::knit_hooks$get("output")

knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$outlines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

options(repos = "http://cran.us.r-project.org")

if (Sys.getenv("USERNAME") == "GHA") {
  googledrive::drive_auth(path = Sys.getenv("GOOGLE_APP_CREDENTIALS"))
}

qmdPath <- normalizePath(getwd(), winslash = "/")

if (!interactive()) { 
  dir.create("~/tutos/LandRExample", recursive = TRUE)
  knitr::opts_knit$set(root.dir = "~/tutos/LandRExample")
}
```

[LandR](https://landr-manual.predictiveecology.org/) is a forest
landscape model implemented as a collection of `SpaDES` modules in R. It
is a reimplementation of LANDIS-II Biomass Succession Extension v.3.2.1,
which at its core is very similar to v7. See the [LandR
Manual](https://landr-manual.predictiveecology.org/), @barros2023 and
@scheller2015 for full details about forest dynamics simulated in LandR.

LandR fully open-source and users are expected to use, modify it and
expand it (e.g. by creating new modules) as they see fit, as long has
modifications are adequately reported. We hope that new modules are
shared with others in the LandR community of users so that all can
benefit.

Each LandR module is hosted in its own GitHub repository. When using and
developing LandR modules, note that modules should be semi-independent.
This is, they should be able to run error-free on their own, even they
don't produce useful outputs in this way. A good example are the "data
preparation" and "data calibration" modules *Biomass_speciesData* and
*Biomass_borealDataPrep* which can run on their own but will not produce
any forest landscape simulations, only the input objects and parameters
that *Biomass_core* (the "simulation module") needs.

In this example, we will setup the workflow published in @barros2023
using `setupProject` from the
[`SpaDES.project`](https://spades-project.predictiveecology.org/)
package and current versions of the LandR modules.



```{r pkgSetup}
#| message: false
#| warning: false
#| results: hide
#| cache: false 

repos <- c("predictiveecology.r-universe.dev", getOption("repos"))
install.packages(c("remotes"), repos = repos)
remotes::install_github("PredictiveEcology/SpaDES.project@transition")
library(SpaDES.project)

# devtools::load_all("../SpaDES.project/")

out <- setupProject(
  ## INPUT OBJECTS -----------------------------------
  sppEquivCol = "Boreal",
  vegLeadingProportion = 0,
  successionTimestep = 10L,
  eventCaching = c(".inputObjects", "init"),
  useParallel = FALSE,
  studyAreaLarge = {
    originalcrs <- "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
    Biomass_corecrs <- "+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"
    largeExtent <- c(xmin = -104.757, xmax = -104.2197, ymin = 55.68663, ymax = 56.20319)
    
    studyAreaL <- terra::vect(terra::ext(largeExtent))
    studyAreaL <-  terra::vect(terra::geom(studyAreaL), "polygons", crs = originalcrs, atts = data.frame(id = 1:length(studyAreaL)))
    studyAreaL <- terra::project(studyAreaL, Biomass_corecrs)
    studyAreaL
  },
  studyArea = {
    originalcrs <- "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
    Biomass_corecrs <- "+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"
    
    smallExtent <- c(xmin = -104.757, xmax = -104.48835, ymin = 55.68663, ymax = 55.94491)
    studyAreaS <- terra::vect(terra::ext(smallExtent))
    studyAreaS <-  terra::vect(terra::geom(studyAreaS), "polygons", crs = originalcrs, atts = data.frame(id = 1:length(studyAreaS)))
    studyAreaS <- terra::project(studyAreaS, Biomass_corecrs)
    studyAreaS
  },
  ## species lists
  sppEquiv = {
    data("sppEquivalencies_CA", package = "LandR")
    sppEquivalencies_CA[grep("Pin", LandR), `:=`(EN_generic_short = "Pine",
                                                 EN_generic_full = "Pine",
                                                 Leading = "Pine leading")]
    sppEquivalencies_CA[grep("Betu_pap", LandR), `:=`(EN_generic_short = "Birch",
                                                      EN_generic_full = "Birch",
                                                      Leading = "Birch leading")]
    ## all Popu will be merged
    sppEquivalencies_CA[grep("Popu_", LandR), `:=`(EN_generic_short = "Poplar",
                                                   EN_generic_full = "Poplar",
                                                   Leading = "Poplar leading")]
    sppEquivalencies_CA[grep("Popu_", LandR), Boreal := "Popu_Spp"]
    
    ## define spp column to use for model
    # sppEquivCol <- "Boreal"
    sppEquivalencies_CA <- sppEquivalencies_CA[!"", on = sppEquivCol]
    sppEquivalencies_CA <- na.omit(sppEquivalencies_CA, sppEquivCol)
    sppEquivalencies_CA
  },
  sppColorVect = {
    LandR::sppColors(sppEquiv, sppEquivCol, newVals = "Mixed", palette = "Accent")
  },
  speciesParams = {
    list(
      "shadetolerance" = list(
        Betu_Pap = 1
        , Lari_Lar = 1
        , Pice_Gla = 2
        , Pice_Mar = 3
        , Pinu_Ban = 1.5
        , Popu_Spp = 1
      )
    )
  },
  paths = list("packagePath" = "packages/",
               "projectPath" = "~/tutos/LandRDemo_coreVeg"),
  packages = c(
    "PredictiveEcology/SpaDES.experiment@development (HEAD)",
    "terra"  ## needed for studyArea, but don't load.
  ),
  options = list(
    "reproducible.destinationPath" = paths$inputPath,
    "spades.inputPath" = paths$inputPath
  ),
    modules = c(
    "CeresBarros/Biomass_speciesData@master",
    "CeresBarros/Biomass_borealDataPrep@master",
    "CeresBarros/Biomass_speciesParameters@master",
    "CeresBarros/Biomass_core@master",
    "CeresBarros/Biomass_validationKNN@master"
  ),
  ## SIMULATION SETUP ------------------------------------
  times = list(start = 2001, end = 2031),
  params = c("PredictiveEcology/PredictiveEcology.org@main/tutos/LandRExample/params.R")
)
```

If you have a look at @sec-castorSetup in the @sec-castorEg chapter, you
will see some differences in the way we setup the two workflows:

-   `paths`. Here we leave the defaults for all paths (see
    [`?setupPaths()`](https://spades-project.predictiveecology.org/reference/setup.html)
    for the list of path options) except where the workflow will be set
    up (`projectPath`) and the location of the package library
    (`packagePath`, which will be placed in `projectPath`).

-   `options`. We also set a couple of "global options" that determine
    the where data will be downloaded to
    ("reproducible.destinationPath"), which will be the same as the
    default directory to look for inputs ("spades.inputPath"). Notice
    how we used `paths$` to get these directory paths from the `paths`
    object that `setupProject` creates (based on the `paths` argument
    above) prior to setting the options.

-   other arguments (`...`). Almost all other arguments in the call
    above were passed to `...`. `studyArea` and `studyAreaL` are two
    polygon shapefiles that *must* be provided by the user. To avoid
    creating them on the `.GlobalEnv` first, we take advantage of
    `setupProject`'s ability to run the code in `{ }` and make these
    polygons. The other object have defaults, but we want to pass
    different values like different species lists (`sppEquiv`) and trait
    values (`speciesParams`).

```{r runSim}
#| message: false
#| warning: false
#| results: hide
#| cache: false 

## remove validation module, can only be run after simulation's done
out$modules <- out$modules[out$modules != "Biomass_validationKNN"]  

## initialise simulation and cache
simInitOut <- do.call(SpaDES.core::simInit, out) |>
  reproducible::Cache()

simOut <- SpaDES.core::spades(simInitOut)
```
