---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
---

# `setupProject` with simple modules

```{r setup}
#| include: false
knitr::opts_chunk$set(tidy = FALSE, 
                      eval = FALSE,
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE)

## truncate text output
hook_output <- knitr::knit_hooks$get("output")

knitr::knit_hooks$set(output = function(y, options) {
  if (!is.null(n <- options$outlines)) {
    y <- xfun::split_lines(y)
    if (length(y) > n) {
      # truncate the output
      y <- c(head(y, n), "....\n")
    }
    y <- paste(y, collapse = "\n")
  }
  hook_output(y, options)
})

options(repos = "http://cran.us.r-project.org")
if (!require("Require")) {
  install.packages("Require", repos = c("https://predictiveecology.r-universe.dev", getOption("repos")))
}

```

`SpaDES.project::setupProject` deals with a number of challenges that crop up with working with multiple modules created by potentially many people.

```{r setupProject, eval=FALSE}
getOrUpdatePkg <- function(p, minVer = "0") {
  if (!isFALSE(try(packageVersion(p) < minVer, silent = TRUE) )) {
    repo <- c("predictiveecology.r-universe.dev", getOption("repos"))
    install.packages(p, repos = repo)
  }
}

getOrUpdatePkg("remotes")

remotes::install_github("PredictiveEcology/Require@simplify4")

getOrUpdatePkg("SpaDES.project", "0.0.8.9049")
```

This example, taken from [Tati Micheletti](https://github.com/tati-micheletti/Micheletti_SpaDES/blob/main/demo.R) shows how to get several modules and run them together.

`setupProject` does a series of very specific things, in a particular order, so that the chances of package conflicts and user errors are minimized. 

```{r setupProject2, eval=FALSE}
##################### PART II: Download the modules and install the needed packages

Setup <- SpaDES.project::setupProject(
  
  paths = list(projectPath = "integratingSpaDESmodules"),
  
  modules = c("tati-micheletti/speciesAbundance@main",
              "tati-micheletti/temperature@main",
              "tati-micheletti/speciesAbundTempLM@main"),
  
  times = list(start = 2013,
               end = 2032),
  
  Restart = TRUE
)
```



```{r runSimInit, eval=FALSE}
##################### PART III: Run SpaDES

results <- do.call(SpaDES.core::simInitAndSpades, Setup)

```


```{r evaluteResults, eval=FALSE}
SpaDES.core::completed(results)
```
