---
title       : "simList -- What is in it?"
author      : "Eliot McIntire"
date        : "May 2024"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, eval = FALSE)
Require::Require("SpaDES.core")
```

## The `simList` {#sec-simlist}

In `R`, there is a default environment, called the `.GlobalEnv`. When
you start typing in R and you create an object, it is "put" in the
`.GlobalEnv`

```{r}
# this is in the .GlobalEnv
a <- 1
environment()
```

This `.GlobalEnv` is convenient to begin in R, but it quickly becomes
limiting. We need more places to put objects, and we need a way to be
more in control of the things that we want to keep, share, modify etc.
In `SpaDES`, this place is called the `simList`. It is a rich data
object that acts like a list or an environment, i.e., we can put objects
into it and use them later.

```{r simListCreate, eval = FALSE}
sim <- simInit()
sim$a <- 1
sim$b <- sim$a + 2
sim$b
envir(sim)
```

```{r simListCreate2, message = FALSE, echo = FALSE}
sim <- SpaDES.core::simInit()
sim$a <- 1
sim$b <- sim$a + 2
sim$b
envir(sim)
```

## Where does `simList` come from?

There are 2 main ways in `SpaDES.core` to create a `simList`:

1.  using [`simInit()`](https://rdrr.io/cran/SpaDES.core/man/simInit.html) creates a `simList` object.
2.  using [`spades()`](https://rdrr.io/cran/SpaDES.core/man/spades.html) , which returns a modified `simList` object.

## How can we specify a `simList`?

A `simList` object is simply a structured data type containing various
elements.

The main components of a `simList` are:

1.  A list of modules used;
2.  The event queue;
3.  A description of the data (object) dependencies.

## How `SpaDES` works: simLists

We can examine the `simList` object structure in the usual R fashion by
printing (`show`ing) it, or by using `str()`:

```{r simlist-structure, echo=TRUE, eval=FALSE}
emptySim <- simInit()
emptySim  # same as show(emptySim)
str(emptySim)
```

**NOTE:** `simList`s are S4 objects, so we can use `getSlots()` and
`slotNames()` to examine the object.

See also [`?'.simList-class'`](https://rdrr.io/cran/SpaDES.core/man/simList-class.html)

## Accessing the parts of a simLists

::: columns-2
```{r simList-accessors, echo=FALSE, eval=TRUE}
emptySim <- SpaDES.core::simInit()
slots <- paste0(slotNames(emptySim))
accessors <- c("modules", "params", "events", "current", "completed", "depends",
               "times", "inputs", "outputs", "paths", "envir", "objects") |>
  paste0("()")
data.frame(slot = slots, accessor = accessors) |>
  knitr::kable(format = "html", padding = 0)
```

```{r simList-accessors2, echo=FALSE, eval=TRUE}
other_accessors <- c("packages", "globals", "start", "end", "timeunit", "timeunits",
                     "paths", "cachePath", "inputPath", "outputPath", "modulePath") |>
  paste0("()")
data.frame(other_accessors = other_accessors) |>
  knitr::kable(format = "html", padding = 0)
```
:::

## Model specification

Simple examples (using demo modules) of `simInit()` and `spades()`
calls.

-   examine `simList` structure before and after the `spades()` call

### Initiliaze an empty `simList` object

```{r empty-simList, eval = FALSE}
emptySim <- simInit()
```

```{r empty-simList2, echo = FALSE}
emptySim <- SpaDES.core::simInit()
```

### Inspect the empty `simList` object

1.  What are the names of each of the slots in the simList?
2.  What do each of these slots contain? Hint: see
    [`?.simList-class`](https://rdocumentation.org/packages/SpaDES.core/versions/2.0.5/topics/simList-class)

```{r inspect-emptySim}
str(emptySim)
slotNames(emptySim)
getSlots('simList')
?'.simList-class'
```

### Inspect the new `simList` object

Use the `simList` accessor functions to determine the following:

1.  the modules included in the simulation and their package
    dependencies;
2.  the global parameter values used;
3.  the module-specific parameter values used;
4.  the simulation start and end times;
5.  the timeunits used by the modules in the simulation;
6.  the scheduled and completed event queues;
7.  the objects (functions, data, etc.) used in the simulation;
8.  the file paths used for simulation inputs and outputs.

```{r inspect-simList, eval=FALSE}
# list modules used in the simulation
modules(mySim)

# list module dependencies and packages used
depends(mySim)
packages(mySim)

# list global and module-specific param values
globals(mySim)
params(mySim)
P(mySim)       ## bonus: how do params() and P() differ?

# list start and end times
times(mySim)
start(mySim)
end(mySim)

# get the simulation and module timeunits
timeunit(mySim)
timeunits(mySim)

# get the scheduled and completed event queues
events(mySim)
completed(mySim)

# list the objects (functions, data, etc.) stored in the simList
objects(mySim)

# list the file paths used in the simulation
paths(mySim)
inputPath(mySim)
outputPath(mySim)
```

### Run the simulation

```{r run-sim,eval=FALSE}
dev()
mySimOut <- Copy(mySim) ## make a deep copy of the simList
mySimOut <- spades(mySimOut)
```

### Compare the simList before and after

```{r compare-sims,eval=FALSE}
times(mySim)
times(mySimOut)

objects(mySim)
objects(mySimOut)
```

## See also

[`?.simList-class`](https://rdocumentation.org/packages/SpaDES.core/versions/2.0.5/topics/simList-class)
[`?simInit`](https://rdocumentation.org/packages/SpaDES.core/versions/2.0.5/topics/simInit)
[`?spades`](https://rdocumentation.org/packages/SpaDES.core/versions/2.0.5/topics/spades)
